---
title: 'Final Stat 184 Project'
author: "Andrew Harley & Abraham Arbelaez"
date: "Due December 17, 2019"
output:  html_notebook
---

# Purpose
The purpose of this project is to figure out how much weather affects the over/under and the outcomes of NFL games over the past 10 years.  

```{r}
library(mosaicData)
library(tidyselect)
library(dplyr)
library(readr)
library(ggplot2)
```


```{r}
spreadspoke_scores <- read_csv("spreadspoke_scores.csv")
nfl_stadiums <- read_csv("nfl_stadiums.csv")
```


```{r}
View(spreadspoke_scores)
View(nfl_stadiums)
```

### Data Wrangling

```{r}
spreadspoke_scores2 <-
spreadspoke_scores %>%
  filter(schedule_season>2009)
```

Since we want data from 2010 on, we filtered out all the cases before then.

```{r}
spreadspoke_scores2 <-
spreadspoke_scores2 %>%
  filter(stadium_neutral== FALSE)
```

Also, we filtered out any games that were at neutral stadiums, so that includes the Super Bowl.

```{r}
spreadspoke_scores2 <-
spreadspoke_scores2 %>%
  mutate(scoretotal = score_away + score_home) %>%
  mutate(over_under_difference = scoretotal - over_under_line) %>%
  mutate(over_or_under = ifelse(over_under_difference > 0, "Over", "Under")) %>%
  filter(over_or_under != "NA")
```

We then created a new variable, which was the total score of each game. Then we used that total score and subtracted it from the over/under line to find the difference between the score and the line. We then created another variable determining if the over or under hit.

```{r}
nfl_stadiums <-
nfl_stadiums %>%
  rename(stadium = stadium_name)
```

In order to join the two data frames together, we renamed the stadium_name to stadium. 


